---
title: "Final Exam"
author: "Alex DeNeve"
date: "December 7, 2017"
output: html_document
---

1.  An ecologist is interested in the intermediate disturbance hypothesis (IDH) 
that postulates that the highest species richness should be found at intermediate levels of 
disturbance. Write a  likelihood function to estimate the  
parameters for a quadratic regression models given in Beckage and Stout 2000 
(http://www.uvm.edu/~bbeckage/Manuscripts/BeckageStout.corrected.2000.jvs.pdf), 
but assuming a Poisson likelihood with a log link function.  Provide your i. *likelihood function and supporting R code for finding the MLE's*, ii. a *plot of the data with the fitted line* and iii. report the *regression parameters (bo, b1, and b2)*, and iv. the value of the *negative log likelihood at the MLEs*.  Note:  You must write your own likelihood function and not use a built in R function or package, e.g, lm, glm, etc.

```{r}
myData<-read.table("http://www.uvm.edu/~bbeckage/Teaching/PBIO_294/Data/idh.csv",sep=',',header=TRUE)
b0<- 0;b1<-1.5;b2<-2.0
parVec<-(b0+b1*myData$nFires+b2*myData$nFires)
parVec

myData
plot(myData)

#theData<- lm(formula=myData$sr~myData$nFires)
#summary(theData)
#b0=y-intercept
#b1=slope4speciesrichness
#b2=slope4#ofFires

```

```{r}
#sr=species richness,nFires=fires in area
#Poisson nll
nllPois<-function(parVec,nFires,sr){
  b0<-parVec[1]
  b1<-parVec[2]
  b2<-parVec[3]
  b0<- -0.7;b1<-.18;b2<-2.0
srPred<- exp(b0+b1*nFires+b2*nFires^2)
#print(srPred)
  nllik<- -sum(dpois(x=sr,lambda=srPred,log=TRUE))
lik<-exp(-nllik)
  return(nllik)
  return(srPred)
}
```


```{r}
parVec=c(0.1,0.1,0.1)
outPois<-optim(par=parVec,fn=nllPois,method="L-BFGS-B",lower=c(-Inf,-Inf), upper=c(Inf,Inf),nFires=myData$nFires,sr=myData$sr)
outPois$par #MLEs at parameters 
outPois$val #log-likelihood at MLEs
outPois$message #errors
myAIC<- 2*3+2*outPois$val
#gives my parameters(b0,b1,b2) use to plot fitted line
```

2. (30 points) Repeat exercise 1 except now fitting the model using either Rstan or Rjags.  You must write your own Stan/JAGS model, meaning that you can not use one of the Stan helper libraries such as the brms or rstanarm libraries. Provide your i. *Stan/JAGS model code and supporting R code for running the Stan/JAGS model*, ii. report the *2.5%, 50%, and 97.5% percentiles for the regression parameters (bo, b1, and b2)*, iii. *plot the posterior distributions for each of these parameters* and iv. *assess the convergence of your chains using the Gelman-Rubin statistic.*

```{r}

#dataLik 
dataLik <- function(param,myData){
    b0 <- param[1]
    b1 <- param[2]
    b2 <- param[3]
    sr<-myData$sr 
    nFires<-myData$nFires
    pFires<-exp(b0+b1*nFires+b2*nFires^2) 
    lik<-dpois(nFires,pFires,log=TRUE)
    logLik <- sum(lik) 
    return(logLik)   
}
myParam<-c(.1,1.5,4)
likelihood <- dataLik(myParam,myData)

#beta0prior
priorB0<-function(b0)                     {
  logLik <- dnorm(0,0.0001,log=TRUE)
  return(logLik)
}
priorB0(b0)

#beta1prior
priorB1<-function(b1)                     {
  logLik <- dnorm(0,0.0001,log=TRUE)
  return(logLik)
}
priorB1(b1)

#beta2prior
priorB2<-function(b2)                     {
  logLik <- dnorm(0,0.0001,log=TRUE)
  return(logLik)
}
priorB2(b2)

#PostB0
postB0<-function(param){
  myData<-read.table("http://www.uvm.edu/~bbeckage/Teaching/PBIO_294/Data/idh.csv",sep=',',header=TRUE)
  return(dataLik(param,myData)+priorB0(b0))
}
param=c(.1,1.5,4)
myPostB0=postB0(param)
myPostB0

#PostB1
postB1<-function(param){
  myData<-read.table("http://www.uvm.edu/~bbeckage/Teaching/PBIO_294/Data/idh.csv",sep=',',header=TRUE)
  return(dataLik(param,myData)+priorB1(b1))
}
param=c(.1,1.5,4)
myPostB1=postB1(param)
myPostB1

#PostB2
postB2<-function(param){
  myData<-read.table("http://www.uvm.edu/~bbeckage/Teaching/PBIO_294/Data/idh.csv",sep=',',header=TRUE)
  return(dataLik(param,myData)+priorB2(b2))
}
param=c(.1,1.5,4)
myPostB2=postB2(param)
myPostB2

#PostLike
postLik<-(likelihood+myPostB0+myPostB1+myPostB2)
postLik
postLik<-function(b0,b2,b3){
  myData<-read.table("http://www.uvm.edu/~bbeckage/Teaching/PBIO_294/Data/idh.csv",sep=',',header=TRUE)
  llik<-(dataLik(c(b0,b1,b2),myData)+priorB0(b0)+priorB1(b1)+priorB2(b2))
  return(llik)
  
}
myPostLik<-postLik(.1,1.5,4)
myPostLik 

max(myData,na.rm=TRUE)
myNewData<-myData[!is.na(myData)]

for(i in 1:nIter)
{
  n1000[i] ~ dpois(parVec[i])
  parVec=exp(b0+b1*myData$nFires+b2*myData$nFires^2)
  log(parVec[i])
  prediction[i] ~ dpois(parVec[i])
}

for(i in 1:nIter)
    if(identical(i %% 1000,0)) cat('iter: ',i,fill=TRUE)
#sampling b0 
    b0Star<-dpois(myNewData,b0[i],log=TRUE)
    r<-min(1,exp(postLik(b0Star,b1[i],b2[i]) - postLik(b0[i],b1[i],b2[i])))
    #could not figure out how to get metropolis step to work, my data was coming out as null if you could send me an email and explain why this error was coming out that would be really helpful, i'd like to figure out exactly how to fix this problem.
#  if(runif(1)<r) {
#      b0[i+1]<-b0Star
#    } else {
#      b0[i+1]<-b0[i]
#    } 
#sampling b1
b1Star<-dpois(myNewData,b1[i],log=TRUE)
    r<-min(1,exp(postLik(b0[i],b1Star,b2[i]) - postLik(b0[i],b1[i],b2[i])))
#if(runif(1)<r) {
#      b1[i+1]<-b1Star
#    } else {
#      b1[i+1]<-b1[i]
#    } 
#samplingb2
    b2Star<-dpois(myNewData,b2[i],log=TRUE)
    r<-min(1, exp( posteriorLik(b0[i],b1[i],b2Star) - posteriorLik(b0[i],b1[i],b2[i])))
#    if(runif(1)<r) {
#      b1[i+1]<-b1Star
#    } else {
#      b1[i+1]<-b1[i]
#    } 
    
```

```{r}
quantile(myPostB0, c(.025,.5,.975))
quantile(myPostB1, c(.025,.5,.975))
quantile(myPostB2, c(.025,.5,.975))
```
```{r}
pred=myPostB0+myPostB1*myData$nFires+myPostB2*myData$nFires^2
plot(myData$sr, pred)
```
```{r}
MCMCdata<-as.mcmc(myData)
plot(MCMCdata)
summary(MCMCdata)
gelman.diag(MCMCdata)
```

3. (40 points) An ecologist expects seed predation to be influenced by the presence of openings (gaps) in the forest overstory and seed mass. The effect of seed mass on predation is expected be quadratic:  Small seeds will be less predated because they offer a lower energetic reward while larger seeds will be too difficult to process, so that medium sized seeds will be most predated.  The effect of a gap is additive with lower predation expected in gaps because seed predators avoid areas of open canopy. The ecologist models the data as a hierarchical logistic regression, assuming that the seed masses are known without observation error.  The model has the form:

Seed lost ~ Binomial(theta) # 1 if seed is predated, 0 if seed is left alone
Logit(theta) = b0i + b1 x gap 
b0i ~ gamma(mu^2/sd^2,mu/sd^2) 
mui = b2 + b3 x seed mass i - b4 x (seed mass i)^2 
sd^2 ~ gamma(0.001,0.001)
b1,b2,b3,b4 ~ normal(0,100)

Fit this model using either Rstan or Rjags.  You must write your own Stan/JAGS model, meaning that you can not use one of the Stan helper libraries such as the brms or rstanarm libraries. Provide your i. *Stan/JAGS model code and supporting R code for running the Stan/JAGS model*, ii. report the *2.5%, 50%, and 97.5% percentiles for the regression parameters (b1, b1, b2, b3, and b4)*, iii. *plot the posterior distributions for each of these parameters* and iv. *assess the convergence of your chains using the Gelman-Rubin statistic.*

The data can be read using:
```{r}
seeds<-read.table("http://www.uvm.edu/~bbeckage/Teaching/PBIO_294/Data/seeds.csv",sep=',',header=TRUE)
Seeds.lost ~ binomial(theta) #1 if seed is predated, 0 if seed is left alone


plot(seeds)
mu ~ dnorm(0,1)
#priors
  priorB0=dnorm(0,100)
  priorB1=dnorm(0,100)
  priorB2=dnorm(0,100)
  priorB3=dnorm(0,100)
  priorB4=dnorm(0,100)
  sdSQ=gamma(0.001)
  param=c(priorB0,priorB1,priorB2,priorB4,sdSQ)
#likelihood
LogLik<-function(myData,param) {
  b0=param[0]
  b1=param[1]
  b2=param[2]
  b3=param[3]
  b4=param[4]
  sdSQ=param[4]
  sdSQ<-gamma(0.001)
  mu=dnorm(0,1)
  b0i=gamma(mu^2/sdSQ)
  log(mu)=b2+b3*seed.mass.i-b4*(seed.mass.i)^2 
  theta=0.1
  logit(theta)=b0i+b1*seeds$gap
  LL <- -sum(dpois(myData$nFires,parVec,log=TRUE))
  return(LL)
}
print(LL)

param=c(priorB0,priorB1,priorB2,priorB4,sdSQ)

#PostB0
B0post<-function(param){
  return(LogLik(param,seeds)+b0)
}
myPostB0=postB0(param)
myPostB0

```







```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

