---
title: "R Notebook"
output: Data Analysis Project
---

Below are the three questions for the PBIO 294 Final Exam.  All work must be your own.  Reference to outside sources should be limited to the documentation for R, Stan, JAGS, or general questions regarding pdf's, etc.  This exam is due by midnight, Monday, 11 Dec 2017. 20% will be deducted for each additonal 6 hours that this exam is turned in late. Please email your R/.Rmd file to brian.beckage@uvm.edu.


An ecologist is interested in the intermediate disturbance hypothesis (IDH) that postulates that the highest species richness should be found at intermediate levels of 
disturbance. You can learn more about the IDH here. 

1. Fit a linear regression model with i) only an intercept, ii) linear response, and  iii) quadratic response using both A) R’s built in lm() and B) the ulam function (i.e., Stan) within the rethinking package. 

2. Evaluate the mcmc chain in the fits from B) above to demonstrate convergence following the guidance that McElreath provides in chapter 9 (e.g., ‘But for typical regression models, you can live by the motto one short chain to debug, four chains for verification and inference.’)

3. Compare model fit using AIC and BIC for A) above and using WAIC and PSIS  for B).  Interpret these metrics.

4. What is your conclusion with respect to IDH?  Explain with support from your model fits.

5. How might you improve your model for these data? (Hint: Is a normal regression most appropriate for these data?)



The data may be read using:
```{r}
srDat<-read.table("https://brianbeckage.github.io/teaching/PBIO_294/Data/idh.csv",sep=',',header=TRUE)
head(srDat)
plot(sr~nFires, data=srDat)

```

```{r}

out <- lm(sr~nFires + I(nFires^2), data=srDat)

plot(sr~nFires, data=srDat)
curve(out$coefficients[1]+out$coefficients[2]*x + out$coefficients[3]*x^2,add=TRUE)


```





