---
title: "10_Ch11_Solutions"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

**11E1. If an event has probability 0.35, what are the log-odds of this
event?**

```{r}
log(0.35/(1-0.35))
```

11E2. If an event has log-odds 3.2, what is the probability of this
event?

log(x/(1-x)) = 3.2

x/(1-x) = exp(3.2)

x = (1-x) \* exp(3.2)

x = exp(3.2) - x\*exp(3.2)

x + x\*exp(3.2) = exp(3.2)

x(1+exp(3.2)) = exp(3.2)

x= exp(3.2)/(1+exp(3.2))

```{r}
exp(3.2)/(1+exp(3.2))

# or using a built-in function in the rethinking library
library(rethinking)
inv_logit(3.2)
```

**11E3. Suppose that a coefficient in a logistic regression has value
1.7. What does this imply about the proportional change in odds of the
outcome?**

```{r}
exp(1.7)
```

This means that each unit change in the predictor variable multiplies
the odds of the event by \~5.5. To demystify this relationship a little,
if the linear model L is the log-odds of the event, then the odds of the
event are just exp(L). Now we want to compare the odds before and after
increasing a predictor by one unit. We want to know how much the odds
increase, as a result of the unit increase in the predictor.

We can use algebra to solve this problem: exp(α + βx) Z = exp(α + β(x +
1))

The left side is the odds of the event, before increasing x. The Z
represents the proportional change in odds that we’re going to solve
for. It’s unknown value will make the left side equal to the right side.
The right side is the odds of the event, after increasing x by 1 unit.
So we just solve for Z now. The answer is Z= exp(β), i.e.,

Z = exp(α + β(x + 1)) / exp(α + βx) Z = exp(α + βx + β) / exp(α + βx) Z
= exp(α + βx)\*exp(β) / exp(α + βx) = exp(B)

And that’s where the formula comes from.

11M3. Explain why the logit link is appropriate for a binomial
generalized linear model.

It is conventional to use a logit link for a binomial GLM because we
need to map the continuous linear model value to a probability parameter
that is bounded between zero and one. The inverse-logit function, often
known as the logistic, is one way (but not the only way, e.g., the
probit link) to do this.

There are deeper reasons for using the logistic. It arises naturally
when working with multinomial probability densities. There was a hint of
this in one of the Overthinking boxes in Chapter 9, in which you saw how
to derive when the binomial distribution has maximum entropy.

11H3. The data contained in data(salamanders) are counts of salamanders
(Plethodon elongatus) from 47 different 49-m2 plots in northern
California.181 The column SALAMAN is the count in each plot, and the
columns PCTCOVER and FORESTAGE are percent of ground cover and age of
trees in the plot, respectively. You will model SALAMAN as a Poisson
variable.

(a) Model the relationship between density and percent cover, using a
    log-link (same as the example in the book and lecture). Use weakly
    informative priors of your choosing. Check the quadratic
    approximation again, by comparing quap to ulam. Then plot the
    expected counts and their 89% interval against percent cover. In
    which ways does the model do a good job? A bad job?

This is a classic case in which the raw predictor variables have an inconveniently large scale.
Standardizing the predictors before fitting will help fit the model.

```{r}
require(rethinking)
data(salamanders)
d <- salamanders
d$C <- standardize(d$PCTCOVER)
d$A <- standardize(d$FORESTAGE)
```

The Poisson model is easy enough. But getting the priors to be sensible will take some simulation.
Consider this model formula:


```{r}
f <- alist(
SALAMAN ~ dpois( lambda ),
log(lambda) <- a + bC*C,
a ~ dnorm(0,1),
bC ~ dnorm(0,1) )
```

Let’s simulate directly from the prior and see what the implied observations look like:
```{r}
N <- 50 # 50 samples from prior
a <- rnorm( N , 0 , 1 )
bC <- rnorm( N , 0 , 1 )
C_seq <- seq( from=-2 , to=2 , length.out=30 )
plot( NULL , xlim=c(-2,2) , ylim=c(0,20) ,
xlab="cover (stanardized)" , ylab="salamanders" )
for ( i in 1:N )
lines( C_seq , exp( a[i] + bC[i]*C_seq ) , col=grau() , lwd=1.5 )
```


You may not have a good prior for how large the average salamander count should be. But it should
not be huge. They are hard to find! This prior isn’t awful—it allows some rare explosive trends, but is
mainly staying within the plausible observation range. Let’s try to calm it down a bit by shrinking the prior for bC:

```{r}
bC <- rnorm( N , 0 , 0.5 )
plot( NULL , xlim=c(-2,2) , ylim=c(0,20) ,
xlab="cover (stanardized)" , ylab="salamanders" )
for ( i in 1:N )
lines( C_seq , exp( a[i] + bC[i]*C_seq ) , col=grau() , lwd=1.5 )
```
That’s better—fewer explosive increases in salamanders.
Now let’s update the prior with the data:

```{r}
f <- alist(
SALAMAN ~ dpois( lambda ),
log(lambda) <- a + bC*C,
a ~ dnorm(0,1),
bC ~ dnorm(0,0.5) )
m1 <- ulam( f , data=d , chains=4 )
precis(m1)
```

There seems to be a positive association with percent cover. Let’s plot the posterior predictions:

```{r}
plot( d$C , d$SALAMAN , col=rangi2 , lwd=2 ,
xlab="cover (standardized)" , ylab="salamanders observed" )
C_seq <- seq( from=-2 , to=2 , length.out=30 )
l <- link( m1 , data=list(C=C_seq) )
lines( C_seq , colMeans( l ) )
shade( apply( l , 2 , PI ) , C_seq )
```


This does seem like a case in which the variance is much greater than the mean—an over-dispersion
situation that we’ll discuss the in the next chapter. Still, it is the very high end of forest cover that
shows some high counts.


(b) Can you improve the model by using the other predictor, FORESTAGE?
    Try any models you think useful. Can you explain why FORESTAGE helps
    or does not help with prediction?

The motivation for this second model is the possibility that forest cover is associated causally
with forest age—older forests are less disturbed and have more cover. In that case, it could be that
forage age is a confound. Let’s inspect a model with both cover and forest age.

    ```{r}
f2 <- alist(
SALAMAN ~ dpois( lambda ),
log(lambda) <- a + bC*C + bA*A,
a ~ dnorm(0,1),
c(bC,bA) ~ dnorm(0,0.5) )
m2 <- ulam( f2 , data=d , chains=4 )
precis(m2)
    ```
Notice that the estimate for bA is now nearly zero, with small interval around it. There isn’t much
association between forest age and salamander density, while also conditioning on percent cover.
Why doesn’t forest age help much? It certainly does improve predictions, in the absence of percent
cover—check for yourself by fitting a model that includes only FORESTAGE as a predictor. If all we
knew was forest age, it would be a good predictor (try it!). But compared to percent cover, forest age
doesn’t help us at all.

