### Binomial Regression Model 

We observed seedling survival in a in a forest.  At each seedlng, we measure an integrated metric of light levels. We want to estimate a logistic regression model for these data to determine the relationship between light levels and seedlng survival.  The model will be of the form:
logit(p)<-beta0 + beta1 * light
seedling ~ binomial(n=1, p)

Repeat steps 1-6 from the Poisson regression model above except for the binomial model.

Please use the following code to simulate data in order to recapture model paramters (beta0 and beta1):

Simulating data for a binomial regression. 

```{r simulate data}
beta0<- -1.0 
beta1<- 3.0
light<-runif(1000)
pVect<-exp(beta0+beta1*light)/(1+exp(beta0+beta1*light))
range(pVect)
seedlingSurv<-rbinom(n=length(light),size=1,prob=pVect)

seedlingSurvData<-data.frame(light,seedlingSurv) # if you want a data frame
plot(light,seedlingSurv)
head(seedlingSurvData)

```

```{r likelihood}
nllBinom<-function(parVec,sdlgs,lght){
  b0 <- parVec[1]
  b1 <- parVec[2]
  seedlsurv<-exp(b0+b1*lght)
  nllik <- -sum(dbinom(x = sdlgs, size = length(lght), log = TRUE))
  return(nllik)
}
```


```{r}
parVec<-c(0.5,1.0) # Initial parameter values 
outBinom<-optim(par=parVec,fn=nllBinom,method="L-BFGS-B",lower=c(-Inf,-Inf),upper=c(Inf,Inf),sdlgs=seedlingSurv,lght=light)
outBinom$par 
outBinom$val 
myAIC<-2*2 + 2*outBinom$val

```

Plotting Binomial regression fit to the data:

```{r}
lightPred<-light[order(light)]
plot(light,seedlingSurv)
lines(lightPred, exp(outBinom$par[1]+outBinom$par[2]*lightPred),col='red',lwd=2)
```

### 2. Compare your maximum likelihood fit to the R glm function.

```{r}
outglm<-glm(seedlingSurv~light, family=binomial,data=seedlingSurvData)
summary(outglm)
```

### 3.  Compare your fit to Stan using brms library

```{r}
library(brms)

outStan <- brm(formula = seedlingSurv ~ light,
            data = seedlingSurvData, family = binomial())

summary(outStan)
plot(outStan)


outStan1 <- brm(formula = seedlingSurv ~ light,
            data = seedlingSurvData, family = binomial(),
            prior = c(set_prior("normal(0,100)", class = "b")),
            warmup = 1000, iter = 2000, chains = 4)

summary(outStan1)
plot(outStan1)


plot(marginal_effects(outStan1), points = TRUE)
```

### 4. Compare your fit to Stan using rstanarm library


```{r}
library(rstanarm)
outrstanarm<-stan_glm(seedlingSurv ~ light,data = seedlingSurvData, family = binomial, iter=2000, warmup=1000, cores=4)

summary(outrstanarm)
plot(outrstanarm)

```

### 5. Fit the Poisson regression using Rstan

```{r}
modelString<-"data {
  int<lower=0> N;
  real<lower=0,upper=1> light[N];
  int<lower=0> nSeedlings[N];
}

parameters {
  real beta0;
  real beta1;
}

transformed parameters {
  real lp[N];
  real <lower=0> mu[N];
  
  for(i in 1:N){
    lp[i] = beta0 + beta1 * light[i];
    mu[i] = exp(lp[i]);
  }
}

model {
  seedlingSurv ~ binomial(prob);
}"
```



```{r}
library(rstan)
myData<-list(seedlingSurv = seedlingSurv,light = light, N = length(light))

resStan <- stan(model_code = modelString, data = myData,
                chains = 3, iter = 3000, warmup = 500, thin = 10)
summary(resStan,par=c('beta0','beta1'))

```


```{r}
library(coda)
post_fit<-As.mcmc.list(resStan) 
plot(post_fit)
```


### 6. Fit the Poisson regression using Rjags


```{r}
modelString<-"
model {
  for (i in 1:N){
    seedlingSurv[i] ~ dbinom(mu[i])
    log(mu[i]) <- beta0 + beta1 * light[i]
  }
  beta0 ~ dnorm(0, .0001)
  beta1 ~ dnorm(0, .0001)
}
"
writeLines(modelString, con='binomReg.jags.txt')
```


```{r}
library(rjags)
jags <- jags.model('binomialReg.jags.txt',
                   data = list('seedlingSurv' = seedlingSurv,
                               'light' = light,
                               'N' = length(light)),
                   inits<-list(
                       list('beta0'=1,'beta1'=2),
                       list('beta0'=1,'beta1'=.2),
                       list('beta0'=.1,'beta1'=2),
                       list('beta0'=2,'beta1'=5)),
                   n.chains = 4,
                   n.adapt = 100)
```


```{r}
update(jags, 1000)

jags.samples(jags,
             c('beta0', 'beta1'),
             10000)
```


```{r}
library(coda)
codaSamples<-coda.samples(jags, c('beta0','beta1'), 10000, 1)
```
```{r}
plot(codaSamples, trace = FALSE, density = TRUE)
summary(codaSamples)
traceplot(codaSamples)
```


