---
title: "Questions Meeting 5: Linear Regression"
output: html_notebook
---

Problem 1: This question uses the !Kung height vs. weight data from the Rethinking package for adults age 18 or older.

Using data from rethinking package. Loading and examining the data.
```{r}
library(rethinking)
data(Howell1)
d <- Howell1
d2 <- d[ d$age >= 18 , ]
head(d2)
plot(height~weight, data=d2)
```


a) Fit a linear regression to these data using the ‘lm’ command within R. The model should predict the observed weight from height and the model should include an intercept and an effect of height. Plot the estimated linear fit superimposed on the data and report the parameter estimates for the intercept, effect of height, and sigma.

```{r}

out<-lm(height~weight,data=d2)
summary(out,correlation=TRUE) # # correlation about -0.99

# Repeating with scaled columns
d3<-d2
d3[c(1,2)] <- lapply(d2[c(1, 2)], function(x) c(scale(x)))
out2<-lm(height~weight,data=d3)
summary(out2,correlation=TRUE) # correlation now 0

plot(height~weight, data=d3)
abline(a=out2$coefficients[1],b=out2$coefficients[2])
library(stats)
sigma(out2)
```


Using rethinking package: normal approximation of posterior
```{r}
library(rethinking)

modelQUAP <- quap(
    alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- intercept + beta*weight ,
        intercept ~ dnorm( 178 , 20 ) ,
        beta ~ dnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
    ) , data=d3 )

precis( modelQUAP )

```
Using rethinking package: MCMC

```{r}

modelMCMC <- ulam(
      alist(
        height ~ dnorm( mu , sigma ) ,
        mu <- intercept + beta*weight ,
        intercept ~ dnorm( 178 , 20 ) ,
        beta ~ dnorm( 0 , 1 ) ,
        sigma ~ dunif( 0 , 50 )
      )
   , data=list(height=d3$height, weight=d3$weight) , chains=3, iter= )

precis(modelMCMC)
```



b) Repeat a) except fitting your model using your own likelihood function rather than R’s built in ‘lm’. Plot the estimated linear fit superimposed on the data and report the parameter estimates for the intercept, effect of height, and sigma. Are they the same as from ‘lm’?

Here is our likelihood function.  We actually work with the negative log likelihood.
Why?

```{r}
nllNormReg<-function(parVec,data=d3){
  wt<-data$weight
  ht<-data$height
  intercept<-parVec[1]
  beta<-parVec[2]
  mySd<-parVec[3]
  htPred<-intercept+beta*wt
  nllik<- -sum(dnorm(x=ht,mean=htPred,sd=mySd,log=TRUE))
  #browser()
  return(nllik)
}
```

Minimizing the negative log likelihood function above.  This is equivalent to finding the maximum likelihood estimate.
```{r}

parVec<-c(1.0,10.0,5.0) # Need some starting parameters for gradient climbing/descending methods
outLM<-optim(par=parVec,fn=nllNormReg,method="L-BFGS-B",
               lower=c(-Inf,-Inf,0.0001),upper=c(Inf,Inf,Inf))
outLM$par # 


```
From R's lm above:
intercept = 5.794e-17 (~ 0)
weight = 7.547e-01
sigma = 0.6569515


Nimble: Fitting lm

```{r}
library(nimble)

myCode <- nimbleCode({
  intercept ~ dnorm(0, sd = 10)
  beta ~ dnorm(0, sd = 10)
  sigma ~ dunif(0, 50)  # prior for variance components based on Gelman (2006)
  for(i in 1:n) {
    mu[i]<-intercept + beta*weight[i]
    height[i] ~ dnorm(mu[i], sd = sigma) 
  }
})

myConstants <- list(n = dim(d3)[1])
myData <- list(weight = d3$weight, height = d3$height)
myInits <- list(intercept = mean(d3$weight), beta = 0, sigma = sqrt(var(d3$weight)))
```


```{r}
mcmc.out <- nimbleMCMC(code = myCode, data = myData, inits = myInits,
                       monitors = c("intercept", "beta","sigma"), 
                       thin = 10,
                       niter = 10000, nburnin=1000,
                       nchains = 1, setSeed = FALSE,
                       constants = myConstants,samplesAsCodaMCMC=TRUE)
```


```{r}
summary(mcmc.out)
```
From R's lm above:
intercept = 5.794e-17 (~ 0)
weight = 7.547e-01
sigma = 0.6569515




Stan: Fitting the same model using Stan

```{r}
library("rstan")
# options(mc.cores = parallel::detectCores())
# rstan_options(auto_write = TRUE)

kung_dat <- list(N=dim(d3)[1],weight = d3$weight,height = d3$height)

fit <- stan(file = 'lm.stan', data = kung_dat,
            iter=10000, warmup=2500,thin=10,chains=4)

print(fit,pars=c('intercept','beta','sigma'),digits_summary = 3,
      probs = c(0.025,0.5,0.975))

```
           mean se_mean    sd   2.5%   50% 97.5% n_eff  Rhat
intercept 0.000   0.001 0.036 -0.071 0.001 0.073  2916 0.999
beta      0.755   0.001 0.035  0.689 0.755 0.824  2791 1.000
sigma     0.659   0.000 0.025  0.612 0.658 0.710  2974 0.999

compared to R's lm above:
intercept = 5.794e-17 (~ 0)
weight = 7.547e-01
sigma = 0.6569515

```{r}
plot(fit)
# pairs(fit,c('intercept','beta','sigma'),las=1)
traceplot(fit)

```
